---
title: "part2"
author: "Erik Johnson"
date: "10/30/2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Overview
Now in the second portion of the project, we're going to analyze the ToothGrowth data in the R datasets package.
## Load the data

```{r message=FALSE,echo=TRUE}
library(dplyr)
library(ggplot2)
```

```{r echo=TRUE}
data("ToothGrowth")

str(ToothGrowth)
```

## Exploratory data analysis

1. Load the ToothGrowth data and perform some basic exploratory data analyses

```{r echo=TRUE}
unique(ToothGrowth$supp)
unique(ToothGrowth$dose)
```

```{r echo=TRUE}
summarizedToothGrowth <-
    ToothGrowth %>%
    group_by(supp, dose) %>% 
    summarize(len.mean = mean(len))

g <- ggplot(summarizedToothGrowth, aes(x = dose, y = len.mean, fill = supp)) + 
    geom_col(position = "dodge") +
    labs(
        x = "Dose",
        y = "Mean Length",
        title = "Average Tooth Growth by Supplement and Dose",
        fill = "Supplement"
    )
print(g)
```

2. Provide a basic summary of the data.

## Confidence intervals and hypothesis tests
3. Use confidence intervals and/or hypothesis tests to compare tooth growth by supp and dose. (Only use the techniques from class, even if there's other approaches worth considering)

```{r echo=TRUE}
len_OJ_05 <- ToothGrowth$len[ToothGrowth$supp == "OJ" & ToothGrowth$dose == 0.5]
len_OJ_10 <- ToothGrowth$len[ToothGrowth$supp == "OJ" & ToothGrowth$dose == 1.0]
len_OJ_20 <- ToothGrowth$len[ToothGrowth$supp == "OJ" & ToothGrowth$dose == 2.0]
len_VC_05 <- ToothGrowth$len[ToothGrowth$supp == "VC" & ToothGrowth$dose == 0.5]
len_VC_10 <- ToothGrowth$len[ToothGrowth$supp == "VC" & ToothGrowth$dose == 1.0]
len_VC_20 <- ToothGrowth$len[ToothGrowth$supp == "VC" & ToothGrowth$dose == 2.0]
len_OJ <- ToothGrowth$len[ToothGrowth$supp == "OJ"]
len_VC <- ToothGrowth$len[ToothGrowth$supp == "VC"]

t.test(len_OJ_05, len_VC_05)
t.test(len_OJ_10, len_VC_10)
t.test(len_OJ_20, len_VC_20)
t.test(len_OJ, len_VC)
```

## Conclusion
4. State your conclusions and the assumptions needed for your conclusions.