---
title: "Placeholder"
output: html_document
---

## Synopsis

Placeholder

## Data Processing

We downloaded the U.S. National Oceanic and Atmospheric Administration's (NOAA) dataset for this assignment from the course link:
[Storm Data](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2) (on 10/23/2017)

The course also provided documentation on the dataset:
* National Weather Service [Storm Data Documentation](https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf)
* National Climatic Data Center Storm Events [FAQ](https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2FNCDC%20Storm%20Events-FAQ%20Page.pdf)

### Reading in the data

We first read in the data from the csv file included in the bz2 archive. The data is a delimited file where fields are delimited with the "," character and missing values are coded as blank fields.

```{r readdata,echo=TRUE}
# Read in the data from the csv file
stormData <- read.csv("StormData.csv", na.strings="")
```

After reading in the data, we check the first few rows (there are 902,297) rows in this dataset.

```{r echo=TRUE}
dim(stormData)
```

```{r echo=TRUE}
head(stormData)
```


###

## Results

### Across the United States, which types of events (as indicated in the EVTYPE variable) are most harmful with respect to population health?

The two variables in the dataset that deal with population health are FATALITIES and INJURIES. 

We start by claculating the average number of fatalities and injuries per event type and put them into the variables Fatalities.Avg and Injuries.Avg, respectively.

We also introduce a new variable, Casualties.Avg, defined as Fatalities.Avg + Injuries.Avg.

```{r avgfatalities,echo=TRUE}
casualties <-
    stormData %>% 
    
    # Group by the event tye
    group_by(EVTYPE) %>% 
    
    # Calculate the average fatalities and injuries for each event type
    summarize(
        Fatalities.Avg=mean(FATALITIES,na.rm=TRUE), 
        Injuries.Avg=mean(INJURIES, na.rm=TRUE)) %>% 
    
    # Add the column for casualties
    mutate(Casualties.Avg = Fatalities.Avg + Injuries.Avg)

dim(casualties)
```

We believe that the most harmful events will be measured primarily by the number of fatalities, and secondarily by the total number of casualties. Therefore, we present the top-10 for each measure, with ties broken first by the number of fatalities and then by the number of injuries.

```{r orderedFatalities,echo=TRUE}
# Show the top 10 most severe events, as measured first by fatalities and then by injuries
head(arrange(casualties, desc(Fatalities.Avg), desc(Injuries.Avg)), n=10)
```

```{r orderedCasualties,echo=TRUE}
# Show the top 10 most severe events, as measured first by casualties, and then by fatalities
head(arrange(casualties, desc(Casualties.Avg), desc(Fatalities.Avg), desc(Injuries.Avg)), n=10)
```


2. Across the United States, which types of events have the greatest economic consequences?

Placeholder
