---
title: "Extreme Heat and Hurricane Weather Events Pose Highest Health and Economic Risk"
output: html_document
---

## Synopsis

In this report, we aim to describe the types of weather events with the greatest impact, both in terms of population health (injuries and fatalities), and in terms of economic consequences (total damage in dollars). Our overall hypothesis is that heat waves cause the most harm to population health, while hurricanes result in the most damage in dollars. To investigate this hypothesis, we used the U.S. National Oceanic and Atmospheric Administration's (NOAA) storm database in our analysis, covering numerous weather events from 1950 through 2011. From these data, we found that, indeed, extreme heat events pose the highest mortality risk (4.36 fatalities on average, N=22), while hurricanes pose the highest economic risk in terms of damage ($1.027 billion on average, N=88).

## Data Processing

We downloaded the U.S. National Oceanic and Atmospheric Administration's (NOAA) dataset for this assignment from the course link:
[Storm Data](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2) (on 10/23/2017). We named the file "StormData.csv.bz2".

The course also provided documentation on the dataset:

* National Weather Service [Storm Data Documentation](https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf)
* National Climatic Data Center Storm Events [FAQ](https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2FNCDC%20Storm%20Events-FAQ%20Page.pdf)

### Reading in the data

We first read in the data from the csv file included in the bz2 archive. The data is a delimited file where fields are delimited with the "," character, quotes are enclosed in double quotes ", and missing values are coded as blank fields.

```{r loadPackages,echo=TRUE}
library(dplyr)
library(ggplot2)
library(gridExtra)
library(RColorBrewer)
```

```{r readData,echo=TRUE,cache=TRUE}
# Read in the data from the csv file
stormData <- read.csv("StormData.csv.bz2", na.strings="", quote="\"")
```

After reading in the data, we check the first few rows (there are 902,297) rows in this dataset.

```{r showDataSummary,echo=TRUE}
dim(stormData)
names(stormData)
head(stormData)
```

### Processing the data

The two columns in the dataset that seem to deal with population health are FATALITIES and INJURIES. We define a new column, Casualties = FATALITIES + INJURIES, which we can use as a measure of the total human health impact of an event.

```{r addCasualtiesColumn,echo=TRUE}
# Add the Casualties column
stormData <- 
    stormData %>%
    mutate(Casualties = FATALITIES + INJURIES)
```

Next, the columns in the dataset that seem to deal with economic consequences are PROPDMG, PROPDMGEXP, CROPDMN, and CROPDMGEXP. We assume these correspond to property damage amount, property damage amount scale (thousands, millions, billions), crop damage amount, and crop damage amount scale (thousands, millions, billions), respectively.

Inspecting PROPDMGEXP and CROPDMGEXP, we find that most non-empty records have values of K, M, or B, corresponding to thousands, millions, and billions, respectively.

```{r summarizeDamageAmountScales,echo=TRUE}
stormData %>% group_by(PROPDMGEXP) %>% summarize(N=n())
stormData %>% group_by(CROPDMGEXP) %>% summarize(N=n())
```

However, it appears that several records are miscoded with other values (or at least, the interpretation of those values is uncertain). We omit those in our analysis.

We define new columns, PropertyDamage.Dollars and CropDamage.Dollars, corresponding to the actual dollar amount of property and crop damage, respectively.

```{r addDamageDollars,echo=TRUE}
# Define a function to calculate the total dollar amount
# given the amount value and the scale factor (K, M, B)
calculateDamageDollarAmount <- function(amount, scale) {
    
    # Convert the scale factor to a character
    scaleValue <- as.character(scale)
    
    # Return the amount, multiplying by the proper scale multiplier
    amount *
        ifelse(scaleValue == "K", 10^3,
            ifelse(scaleValue == "M", 10^6,
                ifelse(scaleValue == "B", 10^9, NA)
            )
        )
}

# Add the damage dollar amount columns
# Make sure to multiply by the proper multiplier depending on the scale
stormData <-
    stormData %>%
    mutate(
        PropertyDamage.Dollars = calculateDamageDollarAmount(PROPDMG, PROPDMGEXP),
        CropDamage.Dollars = calculateDamageDollarAmount(CROPDMG, CROPDMGEXP))
        
```

We then add a new column, TotalDamage.Dollars = PropertyDamage.Dollars + CropDamage.Dollars, which we use as a measure of the total economic consequences.

```{r addTotalDamageDollars,echo=TRUE}
# Define a function to calculate total damage in dollars
calculateTotalDamageDollars <- function (propertyDamageDollars, cropDamageDollars) {
    ifelse(
        is.na(propertyDamageDollars), 
        ifelse(
            is.na(cropDamageDollars), 
            NA,
            cropDamageDollars),
        propertyDamageDollars +
            ifelse(
                is.na(cropDamageDollars),
                0,
                cropDamageDollars
            ))
}

# Add the TotalDamage.Dollars column
stormData <- 
    stormData %>%
    mutate(TotalDamage.Dollars = calculateTotalDamageDollars(PropertyDamage.Dollars, CropDamage.Dollars))
```

Since we're interested in answering questions across the U.S. about event types, we group the data on the EVTYPE column and then calculate summary data for each event type. We calculate the mean for FATALITIES, Casualties, and TotalDamage.Dollars, in order to give an idea of how bad each event type is on average.

```{r summarizedStormData,echo=TRUE}
# Group the storm data by the EVTYPE column
summarizedStormData <-
    stormData %>%
    
    # Group by the event type
    group_by(EVTYPE) %>%

    # Calculate the mean and max values
    summarize(
        N = n(),
        Fatalities.Mean = mean(FATALITIES, na.rm=TRUE),
        Casualties.Mean = mean(Casualties, na.rm=TRUE),
        TotalDamage.Dollars.Mean = mean(TotalDamage.Dollars, na.rm=TRUE))
dim(summarizedStormData)
```

Inspecting the summarized data, we find that many coded event types only appear once or a handful of times.
```{r showEventTypeFrequency,echo=TRUE}
nrow(summarizedStormData[summarizedStormData$N == 1,])
nrow(summarizedStormData[summarizedStormData$N < 5,])
nrow(summarizedStormData[summarizedStormData$N < 10,])
nrow(summarizedStormData[summarizedStormData$N < 15,])
```

On the other hand, we're interested in events that happen with some frequency, so we keep only events that have occurred at least 15 times in the data set.

```{r keepFrequentEvents,echo=TRUE}
minOccurrences <- 15
filteredSummarizedStormData <- filter(summarizedStormData, N >= minOccurrences)
dim(filteredSummarizedStormData)
```

## Results

### 1. Across the United States, which types of events (as indicated in the EVTYPE variable) are most harmful with respect to population health?

In order to show which types of events are most harmful with respect to population health, we plot the top-20 event types ranked in order of most fatalities and most casualties.

While the fatality ranking gives an idea of the deadliest events, the casualty ranking gives an idea of how many people can be impacted by an event.

Events with low ranks for both fatalities and casualties would be considered the most impactful.

```{r plotHealthImpact,fig.width=10,fig.height=6,echo=TRUE}
# Define the number of event types to plot
numberToPlot <- 20

# Select the top 20 events, arranged by our various metrics
eventTypesByFatalitiesMean <- 
    head(arrange(filteredSummarizedStormData, desc(Fatalities.Mean), desc(Casualties.Mean)), n = numberToPlot)
eventTypesByCasualtiesMean <-
    head(arrange(filteredSummarizedStormData, desc(Casualties.Mean), desc(Fatalities.Mean)), n = numberToPlot)

# Sort them by their rank
eventTypesByFatalitiesMean$EVTYPE <- factor(eventTypesByFatalitiesMean$EVTYPE, levels = eventTypesByFatalitiesMean$EVTYPE)
eventTypesByCasualtiesMean$EVTYPE <- factor(eventTypesByCasualtiesMean$EVTYPE, levels = eventTypesByCasualtiesMean$EVTYPE)

# Create the colors to use for each event type
# Get the unique events in our sorted lists
uniqueEventTypes <- unique(c(as.character(eventTypesByFatalitiesMean$EVTYPE), as.character(eventTypesByCasualtiesMean$EVTYPE)))
# Select random colors for these events
set.seed(1235)
eventTypeColors <- sample(colorRampPalette(brewer.pal(5, "Set1"))(length(uniqueEventTypes)))
# Associate the event names with the colors
names(eventTypeColors) <- factor(uniqueEventTypes)
# Create a manual fill using these colors
eventTypeColorScale <- scale_fill_manual(name = "Event Type", values = eventTypeColors)

# Create a plot of events, ranked by fatalities
plotByFatalitiesMean <- 
    ggplot(data = eventTypesByFatalitiesMean, mapping = aes(x = EVTYPE, y = Fatalities.Mean, fill = EVTYPE)) +
    geom_col() +
    coord_flip() + 
    scale_x_discrete(limits=rev(levels(eventTypesByFatalitiesMean$EVTYPE))) +
    geom_text(
        aes(x = EVTYPE, y = Fatalities.Mean / 2, hjust=0, vjust=0.5,
            label = 
                paste(
                    as.character(round(Fatalities.Mean, digits=2)), 
                    paste("(N=", as.character(N), ")", sep="")
                )
            )
        ) + 
    theme(axis.text.y = element_blank(), axis.ticks.y = element_blank()) +
    eventTypeColorScale +
    labs(x = "Event Type", y = "Fatalities (mean)", fill = "Event Type")

# Create a plot of events, ranked by casualties
plotByCasualtiesMean <- 
    ggplot(data = eventTypesByCasualtiesMean, mapping = aes(x = EVTYPE, y = Casualties.Mean, fill = EVTYPE)) +
    geom_col() +
    coord_flip() + 
    scale_x_discrete(limits=rev(levels(eventTypesByCasualtiesMean$EVTYPE))) +
    geom_text(
        aes(x = EVTYPE, y = Casualties.Mean / 2, hjust=0, vjust=0.5,
            label = 
                paste(
                    as.character(round(Casualties.Mean, digits=2)), 
                    paste("(N=", as.character(N), ")", sep="")
                )
            )
        ) + 
    theme(axis.text.y = element_blank(), axis.ticks.y = element_blank()) +
    eventTypeColorScale +
    labs(x = "Event Type", y = "Casualties (mean)")

# Plot the fatality and casualty rankings
grid.arrange(plotByFatalitiesMean, plotByCasualtiesMean, ncol=2, nrow=1, top = "Mean Fatalities and Casualties by Event Type")
```

We see that four events relating to heat rank in the top 10 on both rankings: EXTREME HEAT, HEAT WAVE, HEAT, and EXCESSIVE HEAT. EXTREME HEAT ranks both #1 in fatalities and #2 in casualties.

Additionally, 2 events involving water also rank in the top 10 on both rankings: TSUNAMI (#3 on both), and RIP CURRENTS.

HURRICANE/TYPHOON also makes an appearance in the top 10 in both rankings, and shows as #1 in casualties.

Given the high frequency of heat events (N=22, 74, 767, 1678 for EXTREME HEAT, HEAT WAVE, HEAT, and EXCESSIVE HEAT, respectively), relative to other high impact events such as TSUNAMI (N=20), HURRICANE/TYPHOON (N=88), and RIP CURRENTS (N=304), we conclude that heat events have the largest impact on population health, especially EXTREME HEAT due to its high rate of fatalities and EXCESSIVE HEAT due to its high frequency.

### 2. Across the United States, which types of events have the greatest economic consequences?

In order to show which types of events have the greatest economic consequences, we plot the top-20 event types ranked in order of highest total cost in dollars.

We plot the log of the cost to reduce the separation of the values in the plot.

```{r plotEconomicImpact,fig.height=6,echo=TRUE}
# Define the number of event types to plot
numberToPlot <- 20

# Select the top 20 events, arranged by our various metrics
eventTypesByTotalDamageDollarsMean <- 
    head(arrange(filteredSummarizedStormData, desc(TotalDamage.Dollars.Mean)), n = numberToPlot)

# Sort them by their rank
eventTypesByTotalDamageDollarsMean$EVTYPE <- factor(eventTypesByTotalDamageDollarsMean$EVTYPE, levels = eventTypesByTotalDamageDollarsMean$EVTYPE)

# Create a plot of events, ranked by mean total dollar cost
plotByTotalDamageDollarsMean <- 
    ggplot(data = eventTypesByTotalDamageDollarsMean, mapping = aes(x = EVTYPE, y = log(TotalDamage.Dollars.Mean, base = 10), fill = EVTYPE)) +
    geom_col() +
    coord_flip() + 
    scale_x_discrete(limits=rev(levels(eventTypesByTotalDamageDollarsMean$EVTYPE))) +
    geom_text(
        aes(x = EVTYPE, y = log(TotalDamage.Dollars.Mean, base = 10) / 2, hjust=0, vjust=0.5,
            label = 
                paste(
                    format(TotalDamage.Dollars.Mean, big.mark = ",", scientific = TRUE, digits = 4), 
                    paste("(N=", as.character(N), ")", sep="")
                )
            )
        ) + 
    theme(axis.text.y = element_blank(), axis.ticks.y = element_blank()) +
    labs(x = "Event Type", y = "Log(Mean Total Cost [dollars])", title="Mean Total Cost by Event Type")

# Plot the total cost rankings
plotByTotalDamageDollarsMean
```

We clearly see that hurricanes, storm surge, and floods are by far the most damaging in terms of damage in dollars.

### Session info
We end by showing our session info.
```{r echo=TRUE}
sessionInfo()
```
